\ourTitle

### 2022-06-01 paper/manuscript.tex
arxiv version

%%
%% This is file `sample-lualatex.tex', % generated with the docstrip utility.
%%
%% The original source files were:
%%
%% samples.dtx  (with options: `sigconf')
%% 
%% IMPORTANT NOTICE:
%% 
%% For the copyright see the source file.
%% 
%% Any modified versions of this file must be renamed % with new filenames
%distinct from sample-lualatex.tex.
%% 
%% For distribution of the original source see the terms % for copying and
%modification in the file samples.dtx.
%% 
%% This generated file may be distributed as long as the % original source
%files, as listed above, are part of the % same distribution. (The sources need
%not necessarily be % in the same archive or directory.)
%%
%%
%% Commands for TeXCount TC:macro \cite [option:text,text] TC:macro \citep
%[option:text,text] TC:macro \citet [option:text,text] TC:envir table 0 1
%TC:envir table* 0 1 TC:envir tabular [ignore] word TC:envir displaymath 0 word
%TC:envir math 0 word TC:envir comment 0 0
%%
%%
%% The first command in your LaTeX source must be the \documentclass command.
\documentclass[sigconf, anonymous, review]{acmart}
\usepackage{probs}
\usepackage[linesnumbered,ruled]{algorithm2e}
\mathchardef\mhyphen="2D
%%
%% \BibTeX command to typeset BibTeX logo in the docs
\AtBeginDocument{%
  \providecommand\BibTeX{{%
    \normalfont B\kern-0.5em{\scshape i\kern-0.25em b}\kern-0.8em\TeX}}}

%% Rights management information.  This information is sent to you % when you
%complete the rights form.  These commands have SAMPLE % values in them; it is
%your responsibility as an author to replace % the commands and values with
%those provided to you when you % complete the rights form.
\setcopyright{acmcopyright}
\copyrightyear{2022}
\acmYear{2022}
%\acmDOI{10.1145/1122445.1122456}
\acmConference[KDD'22]{the 28th ACM SIGKDD Conference on Knowledge Discovery and
Data Mining}{August 2022}{Washington, DC, USA}

@@

%%
%% The "title" command has an optional parameter, % allowing the author to
%define a "short title" to be used in page headers.
\title{Predictive State Geometry via Cantor Embeddings and Wasserstein Distance}
%%
%% The "author" command and its associated commands are used to define % the
%authors and their affiliations. % Of note is the shared affiliation of the
%first two authors, and the % "authornote" and "authornotemark" commands % used
%to denote shared contribution to the research.
\author{Samuel P. Loomis}
\email{sloomis@ucdavis.edu}
\orcid{0000-0003-1587-9874}

\author{James P. Crutchfield}
\email{chaos@ucdavis.edu}
\orcid{0000-0003-4466-5410}
\affiliation{%
  \institution{Complexity Sciences Center and Physics Department, \\University of California at Davis}
  \streetaddress{One Shields Ave.}
  \city{Davis}
  \state{CA}
  \country{USA}
  \postcode{95616}
}

%%
%% By default, the full list of authors will be used in the page % headers.
%Often, this list is too long, and will overlap % other information printed in
%the page headers. This command allows % the author to define a more concise
%list % of authors' names for this purpose.
\renewcommand{\shortauthors}{Loomis and Crutchfield}

%%
%% The abstract is a short summary of the work to be presented in the % article.
\begin{abstract}
  Predictive states for stochastic processes are a nonparametric and
  interpretable construct with relevance across a multitude of modeling
  paradigms. Recent progress on the self-supervised reconstruction of predictive
  states from time-series data focused on the use of reproducing kernel
  Hilbert spaces. Here, we examine how Wasserstein distances may be used to
  detect predictive equivalences in symbolic data. We construct a
  finite-dimensional embedding of sequences based on the Cantor set and use
  distances in this embedding to compute Wasserstein distances between
  distributions over sequences (``predictions''). We show that analyzing the
  resulting geometry via hierarchical clustering and dimension reduction
  provides insight into the temporal structure of processes ranging from the
  relatively simple (\emph{e.g.}, finite-state hidden Markov models) to the
  very complex (\emph{e.g.}, infinite-state indexed grammars).
\end{abstract}

%%
%% The code below is generated by the tool at http://dl.acm.org/ccs.cfm. %
%Please copy and paste the code instead of the example below.
%%
\begin{CCSXML}
  <ccs2012>
  <concept>
  <concept_id>10002950.10003648.10003649</concept_id>
  <concept_desc>Mathematics of computing~Probabilistic representations</concept_desc>
  <concept_significance>500</concept_significance>
  </concept>
  <concept>
  <concept_id>10002950.10003648.10003649.10003657</concept_id>
  <concept_desc>Mathematics of computing~Nonparametric representations</concept_desc>
  <concept_significance>500</concept_significance>
  </concept>
  <concept>
  <concept_id>10002950.10003648.10003700.10003701</concept_id>
  <concept_desc>Mathematics of computing~Markov processes</concept_desc>
  <concept_significance>500</concept_significance>
  </concept>
  <concept>
  <concept_id>10002950.10003648.10003700</concept_id>
  <concept_desc>Mathematics of computing~Stochastic processes</concept_desc>
  <concept_significance>500</concept_significance>
  </concept>
  <concept>
  <concept_id>10002950.10003648.10003688.10003693</concept_id>
  <concept_desc>Mathematics of computing~Time series analysis</concept_desc>
  <concept_significance>500</concept_significance>
  </concept>
  <concept>
  <concept_id>10002950.10003741.10003742</concept_id>
  <concept_desc>Mathematics of computing~Topology</concept_desc>
  <concept_significance>300</concept_significance>
  </concept>
  <concept>
  <concept_id>10002950.10003712</concept_id>
  <concept_desc>Mathematics of computing~Information theory</concept_desc>
  <concept_significance>300</concept_significance>
  </concept>
  <concept>
  <concept_id>10003752.10003766</concept_id>
  <concept_desc>Theory of computation~Formal languages and automata theory</concept_desc>
  <concept_significance>300</concept_significance>
  </concept>
  <concept>
  <concept_id>10003752.10003766.10003771</concept_id>
  <concept_desc>Theory of computation~Grammars and context-free languages</concept_desc>
  <concept_significance>300</concept_significance>
  </concept>
  </ccs2012>
\end{CCSXML}

\ccsdesc[500]{Mathematics of computing~Probabilistic representations}
\ccsdesc[500]{Mathematics of computing~Nonparametric representations}
\ccsdesc[500]{Mathematics of computing~Stochastic processes}
\ccsdesc[500]{Mathematics of computing~Time series analysis}
\ccsdesc[300]{Mathematics of computing~Markov processes}
\ccsdesc[300]{Mathematics of computing~Topology}
\ccsdesc[300]{Mathematics of computing~Information theory}
\ccsdesc[300]{Theory of computation~Formal languages and automata theory}
\ccsdesc[300]{Theory of computation~Grammars and context-free languages}

%%
%% Keywords. The author(s) should pick words that accurately describe % the work
%being presented. Separate the keywords with commas.
\keywords{time series, predictive states, Cantor set, Wasserstein distance, hierarchical clustering}

%% A "teaser" image appears between the author and affiliation % information and
%the body of the document, and typically spans the % page.

%%
%% This command processes the author and affiliation and title % information and
%builds the first part of the formatted document.
\maketitle

@@

Predictive states are employed for inference and modeling in dynamical
systems \cite{Crut92c,Emen16a,Salo19a}, renewal processes and spike-trains
\cite{Marz14e,Marz15a}, condensed matter physics \cite{Varn12a,Varn14a}, and
spatiotemporal systems \cite{Rupe19a,Rupe19b,Rupe20a}. A deep mathematical
theory of predictive-state inference has been correspondingly developed
\cite{Jaeg00a,Shal02a,Jame04a,Stil07b,Stre13a,Thon15a,Marz19a,Brod20a}. Despite
this, universal conditions for predictive-state convergence were not given until
recently. It is now known that if the dataset $x_1 \dots x_L$ is drawn from any
stationary process and if $L$ is sufficiently large, then Eq.
\eqref{eq:convergence} converges for any $n$ and a probability-$1$ subset of
pasts $\overleftarrow{x}$ \cite{Loom21a}. In the language of measures,
$P_{\overleftarrow{x}}$ converges \emph{in distribution}, with respect to the
\emph{product topology} of the space of sequences $\mathcal{X}^\mathbb{N}$.

The broad goals of predictive-state analysis are threefold \cite{Shal98a}. The
first is to understand the overall structure of how the predictive states
relate to one another geometrically and, possibly, use this geometry to
classify pasts based on equivalence of their predictive states. The second is
to actually reproduce the prediction to a specified accuracy. The third is to
understand the dynamics of how predictions evolve under a stream of new
observations. The following focuses on the first, as it is a crucial building
block to achieving the other two.

Recent attempts to reconstruct the geometry of predictive states embedded them
in reproducing kernel Hilbert spaces
\cite{Song09a,Song10a,Boot13a,Brod20a,Loom21a}. This was achieved to great
effect largely since convergence in Hilbert spaces generated by universal
kernels is equivalent to convergence in distribution \cite{Srip10a}. That is,
these embeddings allow accurate representations of predictive states because
they respect the product topology of sequences in the same way that predictive
states themselves do.

@@

We refer back frequently to the following example processes:
\begin{enumerate}
\item The \emph{even process} can be generated by repeatedly tossing a coin
	and writing down a $0$ for every tail and $11$ for every head. The process
	is essentially random except that $1$s only appear in contiguous blocks of
	even size bounded by $0$s. The even process has infinite Markov order but
	can be generated by a two-state hidden Markov chain \cite{Crut01a}.
\item The $a^n b^n$ \emph{process} can be generated by choosing a random
	integer $n\geq 1$ (we suppose via a Poisson process) and writing $n$
	$\mathtt{a}$'s followed by an equal number of $\mathtt{b}$'s, and then
	repeating this procedure indefinitely. This results in sequences where any
	contiguous block of $\mathtt{a}$s is followed by a block of $\mathtt{b}$s
	of equal size. The $\mathtt{a}^n \mathtt{b}^n$ process cannot be generated
	by any finite hidden Markov chain, though it is a simple example of a
	probabilistic context-free language \cite{Hopc06a}.
\item The $x+f(x)$ \emph{process} is a probabilistic context-free language
	modeling the syntactic structure of simple mathematical expressions. It has
	terminal symbols $\{\mathtt{(}\ ,\ \mathtt{)}\ ,\ \mathtt{;}\ ,\
	\mathtt{+}\ ,\ \mathtt{f}\ ,\ \mathtt{x}\}$ and nonterminals $\{A,B,C\}$,
	and starts with a sequence of $A$s. Sequences are generated by applying the
	production rules:
\begin{align*}
    A &\mapsto B\ \mathtt{+}\ C\ \mathtt{;}\ |\ C\ \mathtt{;}\\
    B &\mapsto B\ \mathtt{+}\ C\ |\ C\\
    C &\mapsto \mathtt{f(}B\mathtt{)}\ |\ \mathtt{x}
  ~.
\end{align*}
\item The $a^n b^n c^n$ \emph{process} is a probabilistic indexed language
	\cite{Hopc06a} that is analogous to $\mathtt{a}^n \mathtt{b}^n$ except
	after writing the blocks of $\mathtt{a}$'s and $\mathtt{b}$'s, we also
	write a block of $\mathtt{c}$'s of length $n$.
\item The \emph{Morse-Thue process} is generated by sampling from the time
	series of the logistic map at critical ``onset of chaos'' parameter
	$r_c \approx 3.56995$:
\begin{align*}
    y_{t+1} = r y_t(1-y_t)
\end{align*}
	and then coarse-graining the data by taking $x_t = 0$ if $0< y_t \leq
	\frac{1}{2}$ and $x_t = 1$ if $\frac{1}{2} < y_t <1$ \cite{Kurk03a}.
	Alternatively, we can generate this process by starting with a single $0$
	and executing the replacements $0 \mapsto 11$ and $1 \mapsto 01$
	consecutively. The resulting process is an indexed-context free language
	\cite{Crut92c}.
\end{enumerate}

\begin{figure*}[h]
\centering
\includegraphics[width=0.4\linewidth]{../plots/even_CANTOR.png}
\includegraphics[width=0.4\linewidth]{../plots/anbn_CANTOR.png}
\includegraphics[width=0.4\linewidth]{../plots/anbncn_CANTOR.png}
\includegraphics[width=0.4\linewidth]{../plots/x+f(x)_CANTOR.png}
\caption{Cantor plots for the even, $\mathtt{a}^n\mathtt{b}^n$,
	$\mathtt{a}^n\mathtt{b}^n\mathtt{c}^n$, and $\mathtt{x+f(x)}$ processes.
	Each point $(x,y)$ corresponds to a pair of sequences corresponding to the
	past and future, respectively. The symbol on the $x$ ($y$) axis indicates
	that all points above (to the right of) that symbol have a past (future)
	whose most recent observation is that symbol. Though not marked, further
	proportional subdivisions of each segment of the axes indicate the value of
	the second, third, \emph{etc.} symbols. For instance, one can read from the
	$\mathtt{x+f(x)}$ fractal that any past ending in $\mathtt{f}$ must be
	paired with a future beginning in $\mathtt{(f}$ or $\mathtt{(x}$.
	}
\label{fig:cantor}
\Description{Four fractals are displayed, respectively labeled as the even,
	$\mathtt{a}^n\mathtt{b}^n$, $\mathtt{a}^n\mathtt{b}^n\mathtt{c}^n$, and
	$\mathtt{x+f(x)}$ processes.
	}
\end{figure*}

@@

Note that for $|\mathcal{X}|>2$ the embedding $C$ introduces additional
structure that may or may not be desired. Associating each symbol $x$ with an
integer $j_x$ endows an ordinal structure on the set $\mathcal{X}$. This
ordinality is present in the macroscopic geometry of
$C\left(\mathcal{X}^{\mathbb{N}}\right)$. Later, we examine higher-dimension
embeddings that do not have this. However, they come at the cost of increased
computational complexity when determining the Wasserstein distance. So, for now,
we assume that ordinal artifacts are either desired or sufficiently tolerable
as to not outweigh the computational benefits of working in one dimension.

\begin{algorithm}[t]
  \SetKwData{UnqPasts}{UnqPasts}\SetKwData{Cantors}{Cantors}\SetKwData{Wass}{Wass}
  \SetKwFunction{Wasserstein}{Wasserstein}\SetKwFunction{Append}{append}
  \SetKwFunction{Index}{index} \SetKwFunction{Length}{length}
  \SetKwFunction{Matrix}{Matrix}
  \SetKwFunction{CantorWasserstein}{CantorWasserstein}
  \SetKwInOut{Input}{input}\SetKwInOut{Output}{output}

@@

\begin{figure*}[ht]
\centering
\includegraphics[width=0.45\linewidth]{../plots/even_HCLUST.png}
\includegraphics[width=0.45\linewidth]{../plots/anbn_HCLUST.png}
\includegraphics[width=0.45\linewidth]{../plots/anbncn_HCLUST.png}
\includegraphics[width=0.45\linewidth]{../plots/x+f(x)_HCLUST.png}
\caption{(Upper left to lower right) Clustered Cantor diagrams of the even,
	$\mathtt{a}^n\mathtt{b}^n$, $\mathtt{a}^n\mathtt{b}^n\mathtt{c}^n$, and
	$\mathtt{x+f(x)}$ processes. Zoom for detail. For each, the vertical axis
	shows all pasts of a given length $k$ along with their hierarchically
	clustered dendrogram. $k=8$ for the even, $\mathtt{a}^n\mathtt{b}^n$, and
	$\mathtt{a}^n\mathtt{b}^n\mathtt{c}^n$ processes and $k=4$ for the
	$\mathtt{x+f(x)}$ process. For present purposes, the coloring threshold
	was chosen to aid visual interpretation. The lines in each row show the
	empirical distribution of Cantor-embedded futures observed following each
	past. As such, the horizontal axis corresponds exactly to the vertical axis
	of Fig. \ref{fig:cantor}.
	}
\label{fig:hclust}
\Description{Four graphs are displayed. Each vertical axis has a list of
	symbolic sequences organized by a hierarchical clustering tree. The
	horizontal axis is a list of individual symbols. The content of each graph
	is rows of fractal-distributed lines.
	}
\end{figure*}

@@

\begin{figure*}[ht]
\centering
\includegraphics[width=0.45\linewidth]{../plots/even_MDS.pdf}
\includegraphics[width=0.45\linewidth]{../plots/anbn_MDS.pdf}
\caption{Scatterplots of the first two MDS coordinates of the reconstructed
	predictive states: (Left) Even process. (Right) $\mathtt{a}^n\mathtt{b}^n$
	process. Clusters colored according to the scheme determined by the
	dendrogram in Fig. \ref{fig:hclust} and the label on each cluster describes
	the pattern that uniquely characterizes the pasts in that cluster.
	}
\label{fig:mds}
\Description{Two scatter plots with colored clusters of points are displayed,
	with labels next to each cluster describing symbolic patterns.
	}
\end{figure*}

@@

\begin{figure*}[h]
\centering
\includegraphics[width=\linewidth]{../plots/feig_MDS.pdf}
\caption{(Left) Scatterplot of the first two MDS coordinates of the
	reconstructed predictive states for the Morse-Thue process, color-coded by
	cluster. (Right) Scatterplot of the corresponding points in the domain of
	the logistic map, plotting for each point both the present value $y_t$ and
	the next value $y_{t+1}$, with the $x=y$ line for reference. Each pair of
	color-coded arrows shows where each cluster maps to under the action of the
	logistic map. The predictively reconstructed clusters thus correspond to
	dynamically similar neighborhoods of the logistic map domain.
	}
\label{fig:mds-feig}
\Description{A rectangular scatter plot of color-coded points is displayed on
	the left, and on the right a square scatterplot of parabolically arranged
	points with a diagonal line cutting through them. In the right plot the
	points are colored with the same scheme as the left, and similarly colored
	arrows point from clusters to the diagonal line or from the diagonal line
	to the next cluster.
	}
\end{figure*}

@@

%% The acknowledgments section is defined using the "acks" environment % (and
%NOT an unnumbered section). This ensures the proper % identification of the
%section in the article metadata, and the % consistent spelling of the heading.
\begin{acks}
We thank Nicolas Brodu, Adam Rupe, Alex Jurgens, David Gier, and Mikhael
Semaan. JPC acknowledges the kind hospitality of the Telluride Science
Research Center, Santa Fe Institute, Institute for Advanced Study at the
University of Amsterdam, and California Institute of Technology for their
hospitality during visits. This material is based upon work supported by, or
in part by, Templeton World Charity Foundation Diverse Intelligences grants TWCF0440 and TWCF0570 and Foundational Questions Institute and Fetzer Franklin Fund grant number FQXI-RFP-CPW-2007 and grants W911NF-18-1-0028 and
W911NF-21-1-0048 from the U.S. Army Research Laboratory and the U.S. Army
Research Office.

\end{acks}

\section{Reproducibility statement}
For the purposes of reproducibility, we provide a 
GitHub repository \cite{repo} containing
the code necessary to generate this manuscript and its figures, including a
notebook for generating the data used for the examples and for building both
static and interactive figures for further exploration.
%%
%% The next two lines define the bibliography style to be used, and % the
%bibliography file.
\bibliographystyle{ACM-Reference-Format}
\bibliography{ref}

>>>>

%%
%% This is file `sample-lualatex.tex', % generated with the docstrip utility.
%%
%% The original source files were:
%%
%% samples.dtx  (with options: `sigconf')
%% 
%% IMPORTANT NOTICE:
%% 
%% For the copyright see the source file.
%% 
%% Any modified versions of this file must be renamed % with new filenames
%distinct from sample-lualatex.tex.
%% 
%% For distribution of the original source see the terms % for copying and
%modification in the file samples.dtx.
%% 
%% This generated file may be distributed as long as the % original source
%files, as listed above, are part of the % same distribution. (The sources need
%not necessarily be % in the same archive or directory.)
%%
%%
%% Commands for TeXCount TC:macro \cite [option:text,text] TC:macro \citep
%[option:text,text] TC:macro \citet [option:text,text] TC:envir table 0 1
%TC:envir table* 0 1 TC:envir tabular [ignore] word TC:envir displaymath 0 word
%TC:envir math 0 word TC:envir comment 0 0
%%
%%
%% The first command in your LaTeX source must be the \documentclass command.
\documentclass[draft,aps,pre,twocolumn,groupaddress,showkeys,nofootinbib,preprintnumbers,floatfix]{revtex4-2}
\usepackage{probs}
\usepackage{dynlearn}
\usepackage[linesnumbered,ruled]{algorithm2e}
\mathchardef\mhyphen="2D

@@

\def\ourTitle{%
%  Analytical and statistical properties
%  of predictive states
Exploring Predictive States via Cantor Embeddings and Wasserstein Distance
}

\def\ourAbstract{%
Predictive states for stochastic processes are a nonparametric and interpretable
construct with relevance across a multitude of modeling paradigms. Recent
progress on the self-supervised reconstruction of predictive states from
time-series data focused on the use of reproducing kernel Hilbert spaces. Here,
we examine how Wasserstein distances may be used to detect predictive
equivalences in symbolic data. We compute Wasserstein distances between
distributions over sequences (``predictions''), using a finite-dimensional
embedding of sequences based on the Cantor for the underlying geometry. We show
that exploratory data analysis using the resulting geometry via hierarchical
clustering and dimension reduction provides insight into the temporal structure
of processes ranging from the relatively simple (\emph{e.g.}, finite-state
hidden Markov models) to the very complex (\emph{e.g.}, infinite-state indexed
grammars).
}

\def\ourKeywords{%
time series, predictive states, Wasserstein distance, hierarchical clustering
}

\hypersetup{
  pdfauthor={Samuel Loomis},
  pdftitle={\ourTitle},
  pdfsubject={\ourAbstract},
  pdfkeywords={\ourKeywords},
  pdfproducer={},
  pdfcreator={}
}

\author{Samuel P. Loomis}
\email{sloomis@ucdavis.edu}

\author{James P. Crutchfield}
\email{chaos@ucdavis.edu}
\affiliation{Complexity Sciences Center and Department of Physics and Astronomy,
University of California at Davis, One Shields Avenue, Davis, CA 95616}

\date{\today}
\bibliographystyle{unsrt}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\title{\ourTitle}

\begin{abstract}
\ourAbstract
\end{abstract}

\keywords{\ourKeywords}

% \pacs{
% 05.45.-a  %  Nonlinear dynamics and nonlinear dynamical systems
% 89.75.Kd  %  Complex Systems: Patterns
% 89.70.+c  %  Information science
% 05.45.Tp  %  Time series analysis
% 02.50.Ey  %  Stochastic processes
% 02.50.-r  %  Probability theory, stochastic processes, and statistics
% 02.50.Ga  %  Markov processes
% 05.20.-y  %  Classical statistical mechanics
% }

\preprint{\arxiv{2108.XXXXX}}

\date{\today}
\maketitle

\tableofcontents
% Clears up font spacing/overlap in ToC
% with sufficient repetitions of formating (e.g., pdflatex).
% Other front matter packages can alter formating constants.
\makeatletter
\let\toc@pre\relax
\let\toc@post\relax
\makeatother

\setstretch{1.1}

\listoffixmes

@@

Predictive states are employed for inference and modeling in dynamical
systems \cite{Crut92c,Emen16a,Salo19a}, renewal processes and spike-trains
\cite{Marz14e,Marz15a}, condensed matter physics \cite{Varn12a,Varn14a}, and
spatiotemporal systems \cite{Rupe19a,Rupe19b,Rupe20a}. A deep mathematical
theory of predictive-state inference has been correspondingly developed
\cite{Uppe97a,Jaeg00a,Shal02a,Jame04a,Stil07b,Stre13a,Thon15a,Marz19a,Brod20a}.
If the dataset $x_1 \dots x_L$ is drawn from any
stationary process and if $L$ is sufficiently large, then Eq.
\eqref{eq:convergence} converges for any $n$ and a probability-$1$ subset of
pasts $\overleftarrow{x}$ \cite{Uppe97a}. Recently, this result has been further
clarified: in the language of measures,
$P_{\overleftarrow{x}}$ converges \emph{in distribution}, with respect to the
\emph{product topology} of the space of sequences $\mathcal{X}^\mathbb{N}$
\cite{Loom21a}.

The broad goals of predictive-state analysis are threefold \cite{Shal98a}. The
first is to understand the overall structure of how the predictive states
relate to one another geometrically and, possibly, use this geometry to
classify pasts based on equivalence of their predictive states. The second is
to actually reproduce the prediction to a specified accuracy. The third is to
understand the dynamics of how predictions evolve under a stream of new
observations.

The following focuses on the first, as it is a crucial building
block to achieving the other two. Recent attempts to reconstruct the geometry of predictive states embedded them
in reproducing kernel Hilbert spaces
\cite{Song09a,Song10a,Boot13a,Brod20a,Loom21a}. This was achieved to great
effect largely since convergence in Hilbert spaces generated by universal
kernels is equivalent to convergence in distribution \cite{Srip10a}. That is,
these embeddings allow accurate representations of predictive states because
they respect the product topology of sequences in the same way that predictive
states themselves do \cite{Loom21a}. In this paper, we will seek to achieve the same results,
using a procedure which can be easily visualized and interpreted, making it
suitable as a method for exploratory data analysis \cite{Tuke62a,Tuke77a}.

@@

We refer back frequently to the following example processes:
\begin{enumerate}
\item The \emph{even process} can be generated by repeatedly tossing a coin
	and writing down a $0$ for every tail and $11$ for every head. The process
	is essentially random except that $1$s only appear in contiguous blocks of
	even size bounded by $0$s. The even process has infinite Markov order but
	can be generated by a two-state hidden Markov chain \cite{Crut01a}. A typical
	example might look like $01100111101100011$.
\item The $a^n b^n$ \emph{process} can be generated by choosing a random
	integer $n\geq 1$ (we suppose via a Poisson process) and writing $n$
	$\mathtt{a}$'s followed by an equal number of $\mathtt{b}$'s, and then
	repeating this procedure indefinitely. This results in sequences where any
	contiguous block of $\mathtt{a}$s is followed by a block of $\mathtt{b}$s
	of equal size. The $\mathtt{a}^n \mathtt{b}^n$ process cannot be generated
	by any finite hidden Markov chain, though it is a simple example of a
	probabilistic context-free language \cite{Hopc06a}. A typical example might
	look like $\mathtt{abaaabbbabaabb}$.
\item The $x+f(x)$ \emph{process} is a probabilistic context-free language
	modeling the syntactic structure of simple mathematical expressions. It has
	terminal symbols $\{\mathtt{(}\ ,\ \mathtt{)}\ ,\ \mathtt{;}\ ,\
	\mathtt{+}\ ,\ \mathtt{f}\ ,\ \mathtt{x}\}$ and nonterminals $\{A,B,C\}$,
	and starts with a sequence of $A$s. Sequences are generated by applying the
	production rules:
\begin{align*}
    A &\mapsto B\ \mathtt{+}\ C\ \mathtt{;}\ |\ C\ \mathtt{;}\\
    B &\mapsto B\ \mathtt{+}\ C\ |\ C\\
    C &\mapsto \mathtt{f(}B\mathtt{)}\ |\ \mathtt{x}
  ~.
\end{align*}
A typical example might
	look like $\mathtt{x+x;f(x+f(x));f(f(x));}$.
\item The $a^n b^n c^n$ \emph{process} is a probabilistic indexed language
	\cite{Hopc06a} that is analogous to $\mathtt{a}^n \mathtt{b}^n$ except
	after writing the blocks of $\mathtt{a}$'s and $\mathtt{b}$'s, we also
	write a block of $\mathtt{c}$'s of length $n$. A typical example might
	look like $\mathtt{abcaaabbbcccabcaabbcc}$.
\item The \emph{Morse-Thue process} is generated by sampling from the time
	series of the logistic map at critical ``onset of chaos'' parameter
	$r_c \approx 3.56995$:
\begin{align*}
    y_{t+1} = r y_t(1-y_t)
\end{align*}
	and then coarse-graining the data by taking $x_t = 0$ if $0< y_t \leq
	\frac{1}{2}$ and $x_t = 1$ if $\frac{1}{2} < y_t <1$ \cite{Kurk03a}.
	Alternatively, we can generate this process by starting with a single $0$
	and executing the replacements $0 \mapsto 11$ and $1 \mapsto 01$
	consecutively. The resulting process is an indexed-context free language
	\cite{Crut92c}. A typical example might look like
	$11011101010111011101110101011101$ (this is the fifth generation of the
	replacement rule starting from $0$).
\end{enumerate}

\begin{figure*}[ht]
\centering
\includegraphics[width=0.4\linewidth]{../plots/even_CANTOR.png}
\includegraphics[width=0.4\linewidth]{../plots/anbn_CANTOR.png}
\includegraphics[width=0.4\linewidth]{../plots/anbncn_CANTOR.png}
\includegraphics[width=0.4\linewidth]{../plots/x+f(x)_CANTOR.png}
\caption{Cantor plots for the even, $\mathtt{a}^n\mathtt{b}^n$,
	$\mathtt{a}^n\mathtt{b}^n\mathtt{c}^n$, and $\mathtt{x+f(x)}$ processes.
	Each point $(x,y)$ corresponds to a pair of sequences corresponding to the
	past and future, respectively. The symbol on the $x$ ($y$) axis indicates
	that all points above (to the right of) that symbol have a past (future)
	whose most recent observation is that symbol. Though not marked, further
	proportional subdivisions of each segment of the axes indicate the value of
	the second, third, \emph{etc.} symbols. For instance, one can read from the
	$\mathtt{x+f(x)}$ fractal that any past ending in $\mathtt{f}$ must be
	paired with a future beginning in $\mathtt{(f}$ or $\mathtt{(x}$.
	}
\label{fig:cantor}
\end{figure*}

@@

Note that for $|\mathcal{X}|>2$ the embedding $C$ introduces additional
structure that may or may not be desired. Associating each symbol $x$ with an
integer $j_x$ endows an ordinal structure on the set $\mathcal{X}$. This
ordinality is present in the macroscopic geometry of
$C\left(\mathcal{X}^{\mathbb{N}}\right)$. Later, we examine higher-dimension
embeddings that do not have this. However, they come at the cost of increased
computational complexity when determining the Wasserstein distance. So, for now,
we assume that ordinal artifacts are either desired or sufficiently tolerable
as to not outweigh the computational benefits of working in one dimension.
\begin{algorithm}[t]
  \SetKwData{UnqPasts}{UnqPasts}\SetKwData{Cantors}{Cantors}\SetKwData{Wass}{Wass}
  \SetKwFunction{Wasserstein}{Wasserstein}\SetKwFunction{Append}{append}
  \SetKwFunction{Index}{index} \SetKwFunction{Length}{length}
  \SetKwFunction{Matrix}{Matrix}
  \SetKwFunction{CantorWasserstein}{CantorWasserstein}
  \SetKwInOut{Input}{input}\SetKwInOut{Output}{output}

@@

\begin{figure*}[ht]
\centering
\includegraphics[width=0.45\linewidth]{../plots/even_HCLUST.png}
\includegraphics[width=0.45\linewidth]{../plots/anbn_HCLUST.png}
\includegraphics[width=0.45\linewidth]{../plots/anbncn_HCLUST.png}
\includegraphics[width=0.45\linewidth]{../plots/x+f(x)_HCLUST.png}
\caption{(Upper left to lower right) Clustered Cantor diagrams of the even,
	$\mathtt{a}^n\mathtt{b}^n$, $\mathtt{a}^n\mathtt{b}^n\mathtt{c}^n$, and
	$\mathtt{x+f(x)}$ processes. Zoom for detail. For each, the vertical axis
	shows all pasts of a given length $k$ along with their hierarchically
	clustered dendrogram. $k=8$ for the even, $\mathtt{a}^n\mathtt{b}^n$, and
	$\mathtt{a}^n\mathtt{b}^n\mathtt{c}^n$ processes and $k=4$ for the
	$\mathtt{x+f(x)}$ process. For present purposes, the coloring threshold
	was chosen to aid visual interpretation. The lines in each row show the
	empirical distribution of Cantor-embedded futures observed following each
	past. As such, the horizontal axis corresponds exactly to the vertical axis
	of Fig. \ref{fig:cantor}.
	}
\label{fig:hclust}
\end{figure*}

@@

\begin{figure*}[ht]
\centering
\includegraphics[width=0.45\linewidth]{../plots/even_MDS.pdf}
\includegraphics[width=0.45\linewidth]{../plots/anbn_MDS.pdf}
\caption{Scatterplots of the first two MDS coordinates of the reconstructed
	predictive states: (Left) Even process. (Right) $\mathtt{a}^n\mathtt{b}^n$
	process. Clusters colored according to the scheme determined by the
	dendrogram in Fig. \ref{fig:hclust} and the label on each cluster describes
	the pattern that uniquely characterizes the pasts in that cluster.
	}
\label{fig:mds}
\end{figure*}

@@

\begin{figure*}[ht]
\centering
\includegraphics[width=\linewidth]{../plots/feig_MDS.pdf}
\caption{(Left) Scatterplot of the first two MDS coordinates of the
	reconstructed predictive states for the Morse-Thue process, color-coded by
	cluster. (Right) Scatterplot of the corresponding points in the domain of
	the logistic map, plotting for each point both the present value $y_t$ and
	the next value $y_{t+1}$, with the $x=y$ line for reference. Each pair of
	color-coded arrows shows where each cluster maps to under the action of the
	logistic map. The predictively reconstructed clusters thus correspond to
	dynamically similar neighborhoods of the logistic map domain.
	}
\label{fig:mds-feig}
\end{figure*}

@@

%% The acknowledgments section is defined using the "acks" environment % (and
%NOT an unnumbered section). This ensures the proper % identification of the
%section in the article metadata, and the % consistent spelling of the heading.
\section*{Acknowledgements}
We thank Nicolas Brodu, Adam Rupe, Alex Jurgens, David Gier, and Mikhael
Semaan. JPC acknowledges the kind hospitality of the Telluride Science
Research Center, Santa Fe Institute, Institute for Advanced Study at the
University of Amsterdam, and California Institute of Technology for their
hospitality during visits. This material is based upon work supported by, or
in part by, Templeton World Charity Foundation Diverse Intelligences grants TWCF0440 and TWCF0570 and Foundational Questions Institute and Fetzer Franklin Fund grant number FQXI-RFP-CPW-2007 and grants W911NF-18-1-0028 and
W911NF-21-1-0048 from the U.S. Army Research Laboratory and the U.S. Army
Research Office.

\section*{Reproducibility statement}
For the purposes of reproducibility, we provide a 
GitHub repository \cite{repo} containing
the code necessary to generate this manuscript and its figures, including a
notebook for generating the data used for the examples and for building both
static and interactive figures for further exploration.
%%
%% The next two lines define the bibliography style to be used, and % the
%bibliography file.
\bibliography{ref}
